---
title: "Peer-Graded Assignment: Course Project"
output: html_document
---

Instructions
-----------------------------------------

Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:
Dataset: Activity monitoring data [52K]
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as 𝙽𝙰)
*date: The date on which the measurement was taken in YYYY-MM-DD format
*interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

Assignment
-----------------------------------------

### Code for reading in the dataset and/or processing the data
Reading in the dataset:
```{r}
dataset <- read.csv("activity.csv",na.strings = "NA")
```

checking first rows of dataset:
```{r}
head(dataset)
```

Summary of dataset:
```{r}
summary(dataset)
```

### Histogram of the total number of steps taken each day

Calculate total number of steps each day:
```{r}
steps_per_day <- aggregate(dataset$steps, list(dataset$date), sum)
names(steps_per_day) <- c("Date", "Steps")
head(steps_per_day)
summary(steps_per_day)
```

Plot histogram:
```{r}
library(ggplot2)
g1 <- ggplot(steps_per_day) + 
    geom_histogram(aes(x= steps_per_day$Steps), bins=20) +
    labs(title ="Total steps per day", x = "Steps", y = "Frequency")
g1
```

### Mean and median number of steps taken each day
Calculate Mean and Median:
```{r}
mean(steps_per_day$Steps, na.rm = TRUE)
median(steps_per_day$Steps, na.rm = TRUE)
```

Double Check:
```{r}
summary(steps_per_day)
```

### Time series plot of the average number of steps taken

Calculate the average number of steps for each 5-minute interval:
```{r}
avg_steps <- aggregate(dataset$steps, list(dataset$interval), mean, na.rm=TRUE)
names(avg_steps) <- c("Interval", "Steps")
```

Plot time series:
```{r}
g <- ggplot(avg_steps, aes(x = avg_steps$Interval, y = avg_steps$Steps))
g + geom_line() +
    labs(title ="Average number of steps taken", x = "Interval", y = "Average number of steps")
```

### The 5-minute interval that, on average, contains the maximum number of steps
```{r}
avg_steps_int <- aggregate(dataset$steps, list(dataset$interval), mean, na.rm=TRUE)
names(avg_steps_int) <- c("Interval", "Steps")
subset(avg_steps_int, Steps == max(avg_steps_int$Steps))
```

### Code to describe and show a strategy for imputing missing data

The mice package in R, helps you imputing missing values with plausible data values. 
These plausible values are drawn from a distribution specifically designed for each missing datapoint.
```{r}
library(mice)
md.pattern(dataset)
```
Usually a safe maximum threshold is 5% of the total for large datasets.
The output tells us that 15264 samples are complete, 2304 samples miss the Step measurement
In other words: We see that Steps is missing 13% of all datapoints.

The mice() function takes care of the imputing process
```{r eval = FALSE}
tempData <- mice(dataset,m=5,maxit=50,meth='pmm',seed=500)
summary(tempData)
```
m=5 refers to the number of imputed datasets. Five is the default value.
meth='pmm' refers to the imputation method. In this case we are using predictive mean matching as imputation method.

Now we can get back the completed dataset using the complete() function. 
```{r}
completedDataset <- complete(tempData,1)
```
The missing values have been replaced with the imputed values.
```{r}
summary(completedDataset)
```

### Histogram of the total number of steps taken each day after missing values are imputed

Calculate total number of steps each day:
```{r}
steps_per_day2 <- aggregate(completedDataset$steps, list(dataset$date), sum)
names(steps_per_day2) <- c("Date", "Steps")
head(steps_per_day2)
summary(steps_per_day2)
```

Plot histogram:
```{r}
g2 <- ggplot(steps_per_day2) + 
    geom_histogram(aes(x= steps_per_day2$Steps), bins = 20) +
    labs(title ="Total steps per day (complete data)", x = "Steps", y = "Frequency")
g2
```

Comparison:
```{r}
library(gridExtra)
grid.arrange(g1,g2, ncol=2)
```

### Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

Identify weekdays and weekend and divide dataset:
```{r}
completedDataset$date <- as.Date(completedDataset$date)
completedDataset$Weekend <- grepl("S.+",weekdays(completedDataset$date))

WE <- subset(completedDataset, Weekend=="TRUE")
WD <- subset(completedDataset, Weekend=="FALSE")
avg_steps_int_wd <- aggregate(WD$steps, list(WD$interval), mean, na.rm=TRUE)
avg_steps_int_we <- aggregate(WE$steps, list(WE$interval), mean, na.rm=TRUE)
names(avg_steps_int_wd) <- c("Interval", "Steps")
names(avg_steps_int_we) <- c("Interval", "Steps")
```

Plot graphs
```{r}

gWE <- ggplot(avg_steps_int_we, aes(x = avg_steps_int_we$Interval, y = avg_steps_int_we$Steps)) + 
    geom_line() + 
    labs(title ="Average steps per interval on Weekends", x = "Interval", y = "Average amount of Steps")

gWD <- ggplot(avg_steps_int_wd, aes(x = avg_steps_int_wd$Interval, y = avg_steps_int_wd$Steps)) + 
    geom_line() + 
    labs(title ="Average steps per interval on Weekdays", x = "Interval", y = "Average amount of Steps")

grid.arrange(gWD,gWE, nrow=2)
```